Capabilities, MMIO and Robust Safety

### 2020-03-31 capability-meetings-notes/10-03/main.tex
MMIO writeup: new formalization of memory + non-determinism

Notice that, for any address in \MMIO, which value is held by the map $m$ as
that address is now irrelevant. Indeed, $m$ is never read nor modified for
addresses in \MMIO. One could choose to enforce that a particular dummy value is
stored in $m$ for these addresses. Instead, we choose to leave them
unconstrained.

\begin{remark}
  Thomas: Additionally, we might want to disallow in the semantics an execution
  with a \X{PC} register pointing to an $\MMIO$ location (I think this makes
  more sense than reading a random instruction through I/O). Armaël: but it
  makes the semantics of all instructions more complicated.
\end{remark}

@@

The simplest way to account for the trace is to directly expose it in a
monolithic fashion. In that case, the state interpretation additionally holds a
resource for the trace.
\[
  \X{state\_interp} \; (r,m,t) :=
    \X{gen\_heap\_ctx}\; r \ast
    \X{gen\_heap\_ctx}\; (m \setminus \MMIO) \ast
    {\color{BrickRed}\tracefull{t}}
\]

@@

We also assume the boot-code to be initially loaded in memory at address
$l_{\X{boot}}$.
%
Then, the specification for the boot code that we have to prove is stated as
follows:

>>>>

\newcommand{\stdin}{\X{in}}
\begin{remark}
As Lars remarked, the read rule introduces non-determinism into our operational
semantics.
This has lead to some complications in the past (the nature of which is still
mysterious to me; n.b. that we use\\ $\I{ wp\_lift\_atomic\_head\_step\_no\_fork}$
in our WP rules, and that one would allow non-determinism).
(one question I -Thomas- had in this respect; does the $\I{Alloc}$ rule in HeapLang
not introduce non-determinism in a similar way? Why is this not problematic?)
Lars' suggestion was to additionally parameterize the operational semantics by
an input-stream $\stdin$.
The updated \textsc{IOLoad} load rule might then look something like this:
\begin{mathpar}
  \inferrule[IOLoad]
  { r[\X{src}] = (p,g,b,e,a) \\ \X{readAllowed}\; p \\ a \in [b,e) \\ a \in \MMIO
  \\ n = \#\{i \mid t[i] = ( \X{IORead},\_\;,\_) \}  }
  { (r, m, t) \xrightarrow{\I{Load} \; \X{dst} \; \X{src}}
    (r[\X{dst} := \stdin[n]], m, t \app (\X{IORead}, a, \stdin[n])) }
\end{mathpar}

When parameterizing our development with multiple values, we could either write
them out as different Coq \texttt{Parameters} where we need them, or store them
as fields in our \texttt{DriverG} typeclass (the latter is the approach taken
in e.g.\ the code-base of ~\cite{sammler-2020}).
\end{remark}

We now have two options when it comes to modeling the memory $m$:
\begin{enumerate}
\item Include the \MMIO addresses into the map. Notice that, for any address in \MMIO, the specific value held by the map $m$ for
that address is now irrelevant. Indeed, $m$ is never read nor modified for
addresses in \MMIO (possible exception: see next remark). One could choose to enforce that a particular dummy value is stored in $m$ for these addresses. Instead, we choose to leave them
unconstrained.
\item Have $m$ model the physical RAM addresses only. This model is closer to
  the conceptual model of state that we wish to implement, since it does not
  require us to talk about the values in the heap that \MMIO locations are
  associated with. In case it is not harder to implement (which I currently do not think
  it is), it is to be preferred.
\end{enumerate}
Both approaches require us to make sure that no points-to chunks $l \mapsto \_$
with $l \in \MMIO$ can be created. In the next section, we assume the second
approach (until someone proves me wrong about it working).

\begin{remark}
  Additionally, we have the option of disallowing an execution
  with a \X{PC} register pointing to a $\MMIO$ location in the semantics.
  First off, consider that this is an edge case; since we always have points-to
  chunks describing the layout of code in memory (and a points-to chunk for
  location $l$ doubles as a proof that $l \not\in \MMIO$), it is trivial to prove that
  this case would not occur in any of our current examples.
  If we make no changes to the current code base, this behavior would be
  allowed, and the following would happen in both cases described above:
  \begin{enumerate}
  \item An unconstrained value is read from the location. Any instruction could
    be executed.
  \item The predicate $\I{MemLocate}$ in $\I{lang.v}$ will make sure
  that the default value $0$ is read and decoded into an instruction when we try
  to execute an \MMIO location.
  This would not cause a conflict with the current WP rules,
  since a points-to chunk is required in the precondition, ensuring that this
  case does not occur.
  Another alternative is to read the encoding of a $\X{NOP}$ instruction, instead of
  just the value $0$.
\end{enumerate}
  If we were to simply disallow execution when the PC points into \MMIO, we would make the semantics of all instructions slightly more complicated, but in a
  uniform way.
  We could simply add $a \not\in \MMIO$ to the predicate $\I{isCorrectPC}$ (or
  add it as an additional predicate), which is used to define failing execution.
  It would again not be a drastic change, since a points-to for the
  address $a$ the PC points to is required by the WP rules regardless, from which we
  can prove that $a \not\in \MMIO$.
\end{remark}

@@

The simplest way to account for the trace is to directly expose it in a
monolithic fashion. In that case, the state interpretation additionally holds a
resource for the trace:

\begin{align*}
    \X{state\_interp} \; (r,m,t) :=& \X{gen\_heap\_ctx}\; r \ast\\
    &\X{gen\_heap\_ctx}\; m \ast \dom(m) \cap \MMIO = \emptyset \ast
    {\color{BrickRed}\tracefull{t}}
\end{align*}

\begin{remark}
For the case where the \MMIO locations are still part of the heap $m$, this
would look as follows:
\[
  \X{state\_interp} \; (r,m,t) :=
    \X{gen\_heap\_ctx}\; r \ast
    \X{gen\_heap\_ctx}\; (m \setminus \MMIO) \ast
    {\color{BrickRed}\tracefull{t}}
\]

Sanity check: we never have allocation of new locations $l$ in our capability machine, but if we
did, that would work in both settings by requiring that $l \not\in \MMIO$ in the
operational semantics rule for \textsc{Alloc}.
\end{remark}

@@

We also assume the capability machine to initially start execution in the boot code  at address $l_{\X{boot}}$.
%
Then, the syntactic specification for the boot code that we have to prove is stated as follows:
